A Formal Model of Checked C

### 2023-12-01 sandp23/review-answer.md

# Overview

We have uploaded the full version of the paper, which includes an
appendix. The final version of the paper will not have the appendix
but instead will cite a supplemental report.

For the reviewers' convenience we have included a diff between the
original and updated paper version at the end of the PDF. We hope this
will make it easier to see the changes made to the main body of the
paper.

# Major questions

## Formalization status of Theorem 4

Theorem 4, the simulation property, is formalized in PLT Redex, but not
proved; rather, we use randomized testing to give confidence that it
is correct. We don't have a pen and paper proof for it (and did
not claim that we did, in the text).

## Intended guarantee of Theorem 4

Simulation is intended to show that the use of fat pointers in
CoreChkC is not more expressive than a language that does not rely on
fat pointers, as is the expectation for the Checked C language.
The compilation procedure sheds light on how Checked C can be
compiled so as to not need fat pointers. The spatial safety property
is guaranteed by Theorem 3, the blame theorem, and is independent of
what simulation is trying to establish.

## Undefinedness in code examples in Section IV

The target language of compilation, CoreC, is an untyped version of
CoreChkC, i.e., without metadata annotations. We did not intend to
design CoreC as a model of C or an intermediate language such as
LLVM IR. Rather, CoreC can be seen as a device for establishing the fact
that the more expressive semantics of CoreChkC (w.r.t the CheckedC
specification) does not require the use of fat pointers.  The
`deref_array` and `strlen` examples in Section IV (Figs 10-11) are
presented in C syntax to make them easier to understand. As such,
performing pointer arithmetic between a address-0 (null) pointer and a
non-zero integer is defined in CoreC even though it would not be in
actual C, as Reviewer C pointed out. To avoid confusion, we have
dropped type annotations in the examples and made them closer to
actual CoreC, including the use of relative bounds rather than
absolute ones.

## Dereferencing of (un)checked pointers in (un)checked regions

The ordering `u<c` is not a typo, it was the text that was wrong. In
particular, the condition in the rules enforces that unchecked
pointers cannot be used in checked regions; the reverse is
acceptable.

Note that the blame theorem states that whenever the program gets
stuck, we can blame unchecked code. This does not necessarily mean we
always step into a stuck state from the unchecked region. Allowing
access to checked pointers in unchecked regions can allow unchecked
code to corrupt the addresses that checked pointers are assigned
to. The program can get stuck when we later access a corrupted checked pointer
in the checked region, but blame still holds, because the state would
not have been corrupted in the first place if there was no unchecked 
code to corrupt checked pointers. The theorem statements have been
tightened to make this apparent (and make them match the Coq model).

## Discrepancies between the paper's calculus and the Coq formalization

We have added to the paper's supplemental report a dedicated section
that explains the differences between the paper's calculus and the Coq
formalization. For example, Coq syntactically distinguishes the `if`
form for potential ntarray widening and the plain `if` form for
everything else. We have also added a fair bit of extra detail to the
metatheory in Section III.

## Details on model-driven random testing

Our model driven testing approach follows the process recommended by
Pałka et al. [19]. We have added a couple of paragraphs to add more
details, and to clarify that this process is manual, not automatic.

# Other changes

We strived to take care of all of the requests made by the reviewers,
which can be see in the diff at the end of the PDF. Some notes about
specified requests below.

## Reviewer A

> clarify relationship between formalization and implementation

We have added a figure to section I that visualizes the relationship
between our contributions. We also added text to the end of 3.1 to
introduce the Coq and Redex models, and to explain some of their
differences and motivation. We have updated the abstract, conclusions,
and elsewhere to be clearer about what has been proved and what has
been tested.

> correct and address mistakes, typos, confusing terminology and code snippets

We have updated all of the examples. The `strcat` example has been
split into two figures, one in section 3 and one in seciton 4, and
renamed `safe_strcat` to make its intended semantics more clear
(thanks for the suggestion). The two examples intend to implement the
same functionaity, but differ according the semantics of bounds
widening; see the discussion at the end of Section IV-B.

> address pressing questions in the paper.

We have addressed the pressing questions as described in the main
response above, We have gone over all of the reviewer's comments and
attempted to address them; see the diff.

## Reviewer B

> Add details in section 5 (see above)

Done (see above).

> Clarify the status of Theorem 4

Done (see above). Per the reviewer's request, we have also added some
discussion about challenges in the proof, compared to Ruef et al. And
we have clarified the handling of null pointers in the formalism, and
some confusing details in the metatheory.

## Reviewer C

> confirmation that blame theorem issue is essentially just a typo.

Yes this was a typo. We fleshed out the metatheory section be more
precise.

> clarify the policy rules wrt accessing (un)checked pointers in
> (un)checked regions

See above.

> remove, or at least point out and justify the discrepancies between the Coq code and the paper, eg concerning Fig2

We added some discussion to the end of section 3.1 about both the Coq
and Redex models, and added a section to the Appendix about this.

### 2023-12-01 sandp23/checkedc-formal/background.tex
update.

\documentclass[conference]{IEEEtran}

\makeatletter
\def\endthebibliography{\def\@noitemerr{\@latex@warning{Empty `thebibliography' environment}}\endlist
}
\makeatother

\IEEEpeerreviewmaketitle

\usepackage{filecontents}
\usepackage[noadjust]{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{ebproof}
\usepackage{todonotes}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{textcomp}
\usepackage{mathpartir}
\usepackage{color}
\usepackage{xcolor}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}

\usepackage{subcaption} \usepackage{epsfig}
\usepackage{listings}
\usepackage[override]{cmtt}
\usepackage{graphicx}
\usepackage{balance}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{url}
\usepackage{xparse}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{chains,fit,shapes,positioning,calc,arrows.meta,shapes.arrows}

\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}

\lstdefinestyle{customc}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  numbers=left,
  xleftmargin=\parindent,
  language=C,
  columns=flexible,      
  showstringspaces=false,
  basicstyle=\small\sffamily,
  otherkeywords={uint},
  literate={{<-}{{$\leftarrow\,$}}2
            {->}{{$\rightarrow\,$}}2
            {<=}{{$\leq\,$}}2},
  numberstyle=\tiny\rmfamily,
  keywordstyle=\bfseries\color{green!40!black},
  commentstyle=\itshape\color{purple!40!black},
  identifierstyle=\color{blue!80!black},
}
\lstset{language=C,style=customc}

\newenvironment{DIFnomarkup}{}{}
\newcommand{\myparagraph}[1]{\textbf{#1}.\xspace}

\newcommand{\code}[1]{\lstinline|#1|}
%\newcommand{\ignore}[1]{{}}
\newcommand{\kwchecked}{\lstinline|\_Checked|}
\newcommand{\kwunchecked}{\lstinline|\_Unchecked|}
\newcommand{\kwdynamiccheck}{\lstinline|\_Dynamic\_check|}

\newcommand{\var}[1]{#1}
\newcommand{\Arrayptr}[1]{\lstinline|_Array_ptr<|{#1}\lstinline|>|}
\newcommand{\ArrayptrT}{\Arrayptr{$\tau$}}
\newcommand{\Ptr}[1]{\lstinline|_Ptr<|{#1}\lstinline|>|}
\newcommand{\PtrName}{\lstinline|_Ptr|}
\newcommand{\PtrT}{\Ptr{$\tau$}}
\newcommand{\Ntarrayptr}[1]{\lstinline|_Nt_array_ptr<|{#1}\lstinline|>|}
\newcommand{\NtarrayptrT}{\Ntarrayptr{$\tau$}}
\newcommand{\Checkedarr}[2]{{#1}~\lstinline|_Checked[|{#2}\lstinline|]|}
\newcommand{\CheckedarrT}[1]{\Checkedarr{$\tau$}{#1}}

\newif\ifsubmit\submitfalse

\ifsubmit
\newcommand{\mwh}[1]{}
\newcommand{\ashe}[1]{}
\newcommand{\dtarditi}[1]{}
\newcommand{\dvh}[1]{}
\newcommand{\leo}[1]{}
\newcommand{\liyi}[1]{}
\newcommand{\yiyun}[1]{}
\else
\newcommand{\mwh}[1]{\textcolor{red}{Mike: #1}}
\newcommand{\dtarditi}[1]{\textcolor{purple}{David: #1}}
\newcommand{\leo}[1]{\textcolor{green!80!blue}{Leo: #1}}
\newcommand{\liyi}[1]{\textbf{\textcolor{orange}{[ Liyi: #1 ]}}}
\newcommand{\yiyun}[1]{\textcolor{cyan}{Yiyun: #1}}
\newcommand{\dvh}[1]{\textcolor{magenta}{DVH: #1}}
\fi

\newcommand{\lang}{\textsc{CoreChkC}\xspace}
\newcommand{\elang}{\textsc{CoreC}\xspace}

\newtheorem{defi}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{prop}{Proposition}

\newcommand{\CoreChkC}{\textsc{CoreChkC}\xspace}
\newcommand{\checkedc}{\text{Checked C}\xspace}

 \newcommand{\rulelab}[1]{{\small \textsc{#1}}}
\newcommand{\steps}{\ensuremath{\longrightarrow}}
\newcommand{\tbool}{\texttt{bool}}
\newcommand{\tsizeof}{\texttt{sizeof}}
\newcommand{\vbool}{\texttt{Bool}}
\newcommand{\verror}{\texttt{Error}}
\newcommand{\vval}{\mathpzc{val}}
\newcommand{\tfixed}{\texttt{fixedp}}
\newcommand{\tnat}{\texttt{nat}}
\newcommand{\tarr}[2]{\texttt{array}~{#1}~{#2}}
\newcommand{\econst}[2]{({#1}){#2}}
\newcommand{\eindex}[2]{{#1}\texttt{[}{#2}\texttt{]}}
\newcommand{\sassign}[4]{{#1} \leftarrow {#3}~{#2}~{#4}}
\newcommand{\ssassign}[3]{{#1} \xleftarrow{#2} {#3}}
\newcommand{\sif}[3]{\texttt{if}~{#1}~{#2}~{#3}}
\newcommand{\sfor}[3]{\texttt{for}~{#1}~{#2}~{#3}}
\newcommand{\scall}[3]{{#1}\leftarrow {#2}~{#3}}
\newcommand{\sseq}[2]{{#1}\,\texttt{;}\,{#2}}
\newcommand{\sinv}[1]{\texttt{inv}~{#1}}
\newcommand{\inst}[3][ ]{\texttt{#2}^{#1}~{#3}}
\newcommand{\insttwo}[4][ ]{\texttt{#2}^{#1}~{#3}~{#4}}
\newcommand{\instthree}[5][ ]{\texttt{#2}^{#1}~{#3}~{#4}~{#5}}
\newcommand{\iskip}[1]{\inst{ID}{#1}}
\newcommand{\inot}[1]{\inst{X}{#1}}
\newcommand{\ictrl}[2]{\insttwo{CU}{#1}{#2}}
\newcommand{\irz}[3][ ]{\insttwo[#1]{RZ}{#2}{#3}}
\newcommand{\isr}[3][ ]{\insttwo[#1]{SR}{#2}{#3}}
\newcommand{\icnot}[2]{\insttwo{CNOT}{#1}{#2}}
\newcommand{\ilshift}[1]{\inst{Lshift}{#1}}
\newcommand{\irshift}[1]{\inst{Rshift}{#1}}
\newcommand{\irev}[1]{\inst{Rev}{#1}}
\newcommand{\iqft}[2][ ]{\inst[#1]{QFT}{#2}}
\newcommand{\ihad}[1]{\inst{H}{#1}}

\newcommand{\iseq}[2]{{#1}\,\texttt{;}\,{#2}}
\newcommand{\inval}[2]{\insttwo{Nval}{#1}{#2}}
\newcommand{\ihval}[3]{\instthree{Hval}{#1}{#2}{#3}}
\newcommand{\iqval}[2]{\insttwo{Qval}{#1}{#2}}

\newcommand{\instr}{\iota}

\newcommand{\kw}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\subtype}[2]{\ensuremath{\vdash{#1}\prec{#2}}}
\newcommand{\estrlen}[1]{\ensuremath{\kw{strlen}({#1})}}
\newcommand{\estrlentext}{\ensuremath{\kw{strlen}}}

\newcommand{\tarray}[3]{\tarrayb{({#1},{#2})}{#3}}
\newcommand{\tarrayb}[2]{\ensuremath{[{#1}~{#2}]}}
\newcommand{\tntarray}[3]{\tntarrayb{({#1},{#2})}{#3}}
\newcommand{\tntarrayb}[2]{\tarrayb{#1}{#2}_{nt}}
\newcommand{\tallarrayb}[2]{\ensuremath{[{#1}~{#2}]_{\kappa}}}
\newcommand{\tallarraybc}[2]{\ensuremath{[{#1}~{#2}]_{\textcolor{cyan}{\kappa}}}}

\newcommand{\tallarray}[3]{\tallarrayb{({#1},{#2})}{#3}}

\newcommand{\tptr}[2]{\ensuremath{\mathtt{ptr}^{#2}~{#1}}}
\newcommand{\tntptr}[4]{\ensuremath{[({#1},{#2})~{#3}]^{#4}_{nt}*}}

\newcommand{\tarrayptr}[4]{{\tptr{\tarray{#1}{#2}{#3}}{#4}}}
\newcommand{\tntarrayptr}[4]{{\tptr{\tntarray{#1}{#2}{#3}}{#4}}}
\newcommand{\tallarrayptr}[4]{{\tptr{\tallarray{#1}{#2}{#3}}{#4}}}

\newcommand{\tgarray}[4]{\ensuremath{\mathtt{#1}~{#2}~{#3}~{#4}}}
\newcommand{\tstruct}[1]{\ensuremath{\kw{struct}~{#1}}}

\newcommand{\evalue}[2]{\ensuremath{{#1}\!:\!{#2}}}

\newcommand{\emalloc}[1]{\ensuremath{\kw{malloc}({#1})}}
\newcommand{\emalloctext}{\ensuremath{\kw{malloc}}}
\newcommand{\ecall}[2]{\ensuremath{{#1}({#2})}}
\newcommand{\ret}[3]{\ensuremath{\kw{ret}({#1},{#2},{#3})}}
\newcommand{\rettext}{\ensuremath{\kw{ret}}}
\newcommand{\ecast}[2]{\ensuremath{\kw{(}{#1}\kw{)}{#2}}}
\newcommand{\edyncast}[2]{\ensuremath{\langle{#1}\rangle{#2}}}
\newcommand{\edcast}[2]{\ensuremath{\kw{(D},{#1}\kw{)}{#2}}}
\newcommand{\elet}[3]{\ensuremath{\kw{let}~#1\, \texttt{=}\, #2~\kw{in}\;{#3}}}
\newcommand{\elettext}{\ensuremath{\kw{let}}}
\newcommand{\ebinop}[2]{\ensuremath{#1 \plus #2}}
\newcommand{\eassign}[2]{\ensuremath{\texttt{*}{#1}\,\texttt{=}\, {#2}}}
\newcommand{\eassignstack}[2]{\ensuremath{{#1}\,\texttt{=}\, {#2}}}
\newcommand{\efield}[2]{\ensuremath{{#1}\kw{\rightarrow}{#2}}}
\newcommand{\estar}[1]{\ensuremath{\texttt{*}{#1}}}
\newcommand{\eamper}[2]{\ensuremath{\kw{\&}{#1}\kw{\rightarrow}{#2}}}
\newcommand{\eunchecked}[1]{\ensuremath{\kw{unchecked}\;{#1}}}
\newcommand{\euncheckedtext}{\ensuremath{\kw{unchecked}}}
\newcommand{\eif}[3]{\ensuremath{\kw{if\;}(#1)\;{#2}\;\kw{else}\;{#3}}}
\newcommand{\eifa}[3]{\ensuremath{\kw{IF\;}(#1)\;{#2}\;\kw{ELSE}\;{#3}}}
\newcommand{\eifatext}{\ensuremath{\kw{IF}}}

\newcommand{\eiftext}{\ensuremath{\kw{if}}}
\newcommand{\ethentext}{\ensuremath{\kw{then}}}
\newcommand{\eelse}{\ensuremath{\kw{else}}}
\newcommand{\ebreak}{\ensuremath{\kw{break}}}
\newcommand{\ebounds}{\ensuremath{\kw{bounds}}}
\newcommand{\enull}{\ensuremath{\kw{null}}}
\newcommand{\hole}{\ensuremath{\Box}}
\newcommand{\defscope}{\sigma}
\newcommand{\wf}[1]{wf \, #1}
\newcommand{\tint}{\ensuremath{\mathtt{int}}}
\newcommand{\efor}{\ensuremath{\mathtt{for}}}
\newcommand{\ewhile}{\ensuremath{\mathtt{while}}}
\newcommand{\etrue}{\ensuremath{\mathtt{true}}}
\newcommand{\efalse}{\ensuremath{\mathtt{false}}}
\newcommand{\edplus}{\ensuremath{\texttt{++}}}
\newcommand{\edminus}{\ensuremath{\texttt{--}}}

\newcommand{\ememcpy}{\ensuremath{\mathtt{memcpy}}}
\newcommand{\ereturn}{\ensuremath{\mathtt{return}}}
\newcommand{\emain}{\ensuremath{\mathtt{main}}}
\newcommand{\estrcat}{\ensuremath{\mathtt{strcat}}}
\newcommand{\estrcatbad}{\ensuremath{\mathtt{strcat\_b}}}
\newcommand{\heap}{\ensuremath{\mathpzc{H}}}
\newcommand{\eret}[3]{\ensuremath{\kw{ret}({#1},{#2},{#3})}}
\newcommand{\erettext}{\ensuremath{\kw{ret}}}

\newcommand{\plus}{\mathbin{\texttt{+}}}

\newcommand{\fv}{\mathit{FV}}
\newcommand{\dom}{\mathit{dom}}

\newcommand{\fm}{\mathit{FM}}
\newcommand{\size}{\mathit{size}}

\newcommand{\cmode}{\texttt{c}}
\newcommand{\umode}{\texttt{u}}
\newcommand{\bvar}{\ensuremath{\beta}}
\newcommand{\mode}{\textit{mode}}

\newcommand{\cextend}[4]{\ensuremath{#4 = ~ \vdash_{extend} #1, #2, #3}}
\newcommand{\fresh}{\ensuremath{\kw{fresh}}}
\newcommand{\echecknull}[3]{\ensuremath{#3 = ~\vdash_{null}#1, #2}}
\newcommand{\echeckboundsdyn}[5]{\ensuremath{#5 = ~ \vdash_{boundsD}#1, #2, #3, #4}}
\newcommand{\echeckbounds}[5]{\ensuremath{#5 = ~ \vdash_{boundsR}#1, #2, #3, #4}}
\newcommand{\echeckboundsw}[5]{\ensuremath{#5 = ~ \vdash_{boundsW}#1, #2, #3, #4}}
\newcommand{\esizeof}[1]{\ensuremath{\kw{sizeof}(#1)}}
\newcommand{\ewidenstrlen}[5]{\ensuremath{#5 = ~ \vdash_{widenstr}#1, #2, #3, #4}}
\newcommand{\ewidenderef}[4]{\ensuremath{#4 = ~ \vdash_{widenderef}#1, #2, #3}}
\newcommand{\eleq}[2]{\ensuremath{#1 \leq #2}}

\newcommand{\rboundle}[2]{\ensuremath{\vdash_{bounds}#1 \leq #2}}

\def\codesize{\normalsize}
\definecolor{programs}{gray}{0.1}

\lstset{
  basicstyle=\ttfamily,  
}

\makeatletter
\lst@CCPutMacro
    \lst@ProcessOther {"22}{\lst@ifupquote \textquotedbl
                                     \else \char34\relax \fi}
    \@empty\z@\@empty
\makeatother

\lstloadlanguages{C}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{A Formal Model of Checked C}

% \ignore{
% \author{\IEEEauthorblockN{1\textsuperscript{st} Leonidas Lampropoulos}
% \IEEEauthorblockA{\textit{Department of Computer Science} \\
% \textit{University of Maryland}\\
% llampro@cs.umd.edu}
% \and
% \IEEEauthorblockN{1\textsuperscript{st} Michael Hicks}
% \IEEEauthorblockA{\textit{Department of Computer Science} \\
% \textit{University of Maryland}\\
% llampro@cs.umd.edu}
% }
% \author{Leonidas Lampropoulos \and
% Michael Hicks}
% }
\maketitle

\begin{abstract}

  We present a formal model of Checked C, a dialect of C
  that aims to enforce spatial memory safety. Our model pays particular
  attention to the semantics of dynamically sized, potentially null-terminated
  arrays.
We formalize this model in Coq, and prove that any spatial memory
  safety errors can be \emph{blamed} on portions of the program
  labeled \emph{unchecked}; this is a Checked C feature that supports
  incremental porting and backward compatibility.
Our model develops an operational semantics that uses fat pointers to
  guarantee spatial safety. However, we formalize a compilation scheme
  that can yield thin pointers, with bounds information managed using
  inserted code. We show that the generated code faithfully simulates
  the original.
Finally, we build an executable version of our model in PLT Redex, 
  and use a custom random generator
  for well-typed and almost-well-typed terms to find
  inconsistencies between our model and the Clang Checked C
  implementation. We find these steps to be a useful way to co-develop a language
  (Checked C is still in development) and a core model of it.
\end{abstract}

\section{Introduction}\label{sec:intros}

The C programming language remains extremely popular despite the
emergence of new, modern languages. Unfortunately, C programs lack
spatial memory safety, which makes them susceptible to a host
of devastating vulnerabilities, including buffer overflows and
out-of-bounds reads/writes. Despite their long history, buffer
overflows and other spatial safety violations are among the most
prevalent and dangerous vulnerabilities on the Internet today \cite{Zeng:2013:SRF:2534766.2534798}.

Several industrial and research efforts---including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012}---have
explored means to compile C programs 
to automatically enforce spatial safety. These
approaches all impose performance overheads that are deemed too high for
use in deployment. Recently, Microsoft introduced \checkedc, an
open-source extension to C with new types and
annotations whose use can ensure a program’s spatial
safety~\cite{Elliott2018}. Importantly, \checkedc supports development that is
incremental and compositional. Code regions (e.g.,
functions or whole files) designated as \emph{checked} are sure to enforce
spatial safety, a property which is preserved via composition with
other checked regions. But not all regions must be checked: Checked
C's annotated \emph{checked pointers} are binary-compatible with legacy pointers, and
may coexist in the same code, which permits a deliberate (and
semi-automated) refactoring process. Parts of the FreeBSD kernel have
been successfully ported to \checkedc~\cite{duanrefactoring}, and overall, performance
overhead seems low enough for practical deployment.

While \checkedc promises to enforce spatial safety, we might wonder
whether its design and implementation deliver on this promise, or even
what ``spatial safety'' means when a program contains both checked and
unchecked code. In prior work, \citet{ruef18checkedc-incr} developed a core
formalization of \checkedc and with it proved that \emph{checked code cannot 
  be blamed}: any spatial safety violation can only be attributed to
code that is not in a checked region. While their work is a good start,
it fails to model important aspects of \checkedc's functionality,
particularly those involving pointers to arrays. In this paper, we
cover this gap, making three main contributions.

\myparagraph{Dynamically bounded and null-terminated arrays}
Our first contribution is a core formalism called \lang, which extends
\citet{ruef18checkedc-incr} with several new features, most notably 
\emph{dynamically bounded arrays} (Section~\ref{sec:formal}). 
Dynamically bounded arrays are those whose size is
known only at run time, as designated by in-scope variables using
dependent types. A pointer's accessible memory is bounded both above
and below, to admit arbitrary pointer arithmetic.

\lang also models
\emph{null-terminated} arrays, a kind of dynamically bounded array whose upper bound 
defines the array's \emph{minimum} length---additional space is available 
up to a null terminator. For example, the \checkedc type
\lstinline{nt_array_ptr<char> p:count(n)} says that
\lstinline{p} has length \emph{at least} \lstinline{n} (excluding the
null terminator), but further capacity is present if \lstinline|p[n]| is not null. \checkedc (and \lang) supports flow-sensitive
\emph{bounds widening}: statements of the form \lstinline|if (*p) | $s$,
where \lstinline{p}'s type is \lstinline{nt_array_ptr<T> count(0)},
typecheck statement $s$ under the assumption that \lstinline{p} has
type \lstinline{nt_array_ptr<T> count(1)}, 
 i.e., one more than it was,
since the character at the current-known length is
non-null. Similarly, the call \lstinline|n = strlen(p)| will widen
\lstinline|p|'s bounds to \lstinline|n|. Subtyping permits treating
null-terminated arrays as normal arrays of the same size (which
does not include, and thereby protects, the null terminator).

We prove, in Coq, a blame theorem for \lang.  As far as we are aware,
ours is the first formalized type system and proof of soundness for
pointers to null-terminated arrays with expandable bounds.

\myparagraph{Sound compilation of checked pointers} Our second
contribution is a formalization of bounds-check insertion for array
accesses (Section~\ref{sec:compilation}). Our operational semantics
annotates each pointer with metadata that describes its bounds, and
the assignment and dereference rules have premises to confirm the
access is in bounds. An obvious compilation scheme (taken by
Cyclone~\cite{Jim2002,GrossmanMJHWC02}, CCured~\cite{Necula2005}, and
earlier works) would be to translate annotated pointers to multi-word
objects: one word for the pointer, and 1-2 words to describe its lower
and upper bounds. Inserted checks reference these bounds. While
convenient, such ``fat'' pointers are expensive, and break backward
binary compatibility with legacy pointers. We formalize \checkedc's
compilation approach, which uses a single machine word for the
pointer, and adds checks involving the declared bounds (e.g., in a
dependent type) or additional stack-allocated \emph{ghost variables}
to accommodate bounds widening. We show that the compiled program
\emph{simulates} the original by mechanizing \lang and the compilation
judgment in PLT Redex~\cite{pltredex}, and use its random testing
feature to give confidence that simulation holds.

As far as we are aware, \lang is the first formalism to cleanly
separate bounds-checking compilation from the core semantics; prior
work merged the two, conflating \emph{meaning} with
\emph{mechanism}~\cite{Feng2006,Condit2007}. In carrying out the 
formalization, we discovered that our compilation approach is 
more expressive than that proposed in the Checked C
specification~\cite{checkedc} (Section~\ref{sec:disc}); we 
would not have discovered this improvement had we not separated checks from 
semantics.

\myparagraph{Model-based randomized testing} Our third
and final contribution is a strategy and implementation of model-based
randomized testing (Section~\ref{sec:evaluation}).  To check the
correctness of our formal model, we compare the behavior between the
existing Clang \checkedc implementation and our own model. This is
done by a conversion tool that converts expressions from \lang into
actual \checkedc code that can be compiled by the Clang \checkedc
compiler. We build a random generator of programs largely based on the
typing rules of \lang and make sure that, both statically and
dynamically, \lang and Clang Checked C are consistent after
conversion.  This helped rapidly prototype the model and uncovered
several issues in the Checked C compiler.

We begin with a review of \checkedc (Section~\ref{sec:overview}),
present our main contributions (Sections~\ref{sec:formal}--\ref{sec:evaluation}),
and conclude with a discussion of
related and future work (Sections~\ref{sec:related}, \ref{sec:conclude}).

@@

This section describes Checked C, which extends C with new pointer
types and annotations that ensure spatial safety. More
details can be found in a prior overview~\cite{Elliott2018} or the
full specification~\cite{checkedc}.
Checked C is implemented as a fork of Clang/LLVM and is freely
available.\footnote{\url{https://github.com/Microsoft/checkedc-clang}}

@@

}

\caption{Parsing a string of UTF16 hex characters in \checkedc}
\label{fig:checkedc-example}
\end{figure}

\subsection{Checked Pointer Types}
\checkedc introduces three varieties of \emph{checked pointer}:
\begin{itemize}
\item \code{ptr<}$T$\code{>} types a pointer that is either null or
  points to a single object of type $T$.
\item \code{array_ptr<}$T$\code{>} types a pointer that is either null
  or points to an array of $T$ objects. The array width is defined
  by a \emph{bounds} expression, discussed below.
\item \code{nt_array_ptr<}$T$\code{>} is like
  \code{array_ptr<}$T$\code{>} except that the bounds expression
  defines the \emph{minimum} array width---additional objects may
  be available past the upper bound, up to a null terminator.
\end{itemize}
A bounds expression used with the latter two pointer types has three
forms:
\begin{itemize}
\item \code{count(}$e$\code{)} where $e$ defines the array's
  length. Thus, if pointer $p$ has bounds \code{count(n)} then the
  accessible memory is in the range $[p,p+$\code{n}$]$. Bounds
  expression $e$ must be side-effect free and may only refer to
  variables whose addresses are not taken, or adjacent \code{struct}
  fields.
\item \code{byte_count(}$e$\code{)} is like \code{count}, but
  expresses arithmetic using bytes, no objects; i.e.,
  \code{count(}$e$\code{)} used for \code{array_ptr<}$T$\code{>} is
  equivalent to \code{byte_count(}$e\times\texttt{sizeof}(T)$\code{)}
\item \code{bounds(}$e_l$,$e_h$\code{)} where $e_l$ and $e_h$ are
  pointers that bound the accessible region $[e_l,e_h)$ (the
  expressions are similarly restricted). Bounds
  \code{count(}$e$\code{)} is shorthand for
  \code{bounds(}$p, p + e$\code{)}. This most general form of bounds
  expression is useful for supporting pointer arithmetic.
\end{itemize}
The \checkedc compiler will instrument loads and stores of checked
pointers to confirm the pointer is non-null, and the access is within
the specified bounds. For pointers $p$ of type
\code{nt_array_ptr<}$T$\code{>}, such a check could spuriously fail if
the index is past $p$'s specified upper bound, but before the null
terminator. To address this problem, Checked C supports \emph{bounds
  widening}. If $p$ has bounds expression
\code{bounds(}$e_l$,$e_h$\code{)} a program may read from (but not
write to) $e_h$; when the compiler notices that a non-null character
is read at the upper bound, it will extend that bound to $e_h+1$.

\subsection{Example}

Fig.~\ref{fig:checkedc-example} gives an example \checkedc
program.\footnote{Ported from the Parson JSON
parser, \url{https://github.com/kgabis/parson}} 
The function \code{parse_utf16_hex} on lines 1-17 takes as its argument
null-terminated pointer \code{s} from which it attempts to read four
characters. These are interpreted as hex digits and converted to an
\code{uint} returned via parameter \code{result}. At first,
\code{s} has no specific bounds annotation, which we can interpret as
\code{count(0)}; this means that \code{s[0]} may be read on line
5. The true branch of the conditional (which extends all the way to
the brace on line 15) is thus typechecked with \code{s} given a
\emph{widened} bound of \code{count(1)}. Likewise, the conditionals on
lines 6-8 each widen it one further; the widened pointer
(\code{s+4}) is returned on success.

>>>>

@@

This section describes Checked C, which extends C with new pointer
types and annotations that ensure spatial safety. Development of
Checked C was initiated by Microsoft Research in 2015 but starting in
late 2021 was forked and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the full specification~\cite{checkedc}.
Checked C is implemented as an extension of Clang/LLVM and the SSDP
fork is freely available at \url{https://github.com/secure-sw-dev}.

@@

% {\captionsetup[lstlisting]{margin = 8 mm}
%   \begin{lstlisting}[xleftmargin=8 mm]
% array_ptr<int> memcpy(int x,
%      array_ptr<int> a : count(x),
%      array_ptr<int> b : count(x)) : count(x) {
%  for(int i = 0; i < x; ++i) {
%     (*a = *b); a++; b++;
%  }
%  return a;
}

% static int idx = 0;
% int buf (int * reg,
%           nt_array_ptr<int> x,
%           nt_array_ptr<int> y) {
%   if (idx < n)
%     unchecked { * x = &(reg + idx); idx++; }
%   else bug();
%   memcpy(n,x,y);
%   return 0;
% }
% }
\caption{Parsing a String of UTF16 Hex Characters in \checkedc}
\label{fig:checkedc-example}
\end{figure}

% Fig.~\ref{fig:checkedc-example}, all dereferences of pointer \code{a}
% are within the range $[$\code{0}$,$\code{x}$)$ and therefore
% safe. Moreover, if we were to remove the condition \code{i<x} from the
% \code{for} loop, attempting to dereference an out-of-bounds memory
% location, the inserted checks guarantee that the system will raise an
% out-of-bound exception instead of crashing with a segmentation fault
% or, worse, allowing the access to
% happen. %if a out-of-bound error happens, the system causes an out-of-bound exception.

% To insert such checks, both array and NT-array pointers are associated
% with two values, their \emph{bounds}, defining the range of memory
% referenced by the pointers. An array pointer $p$ with
% bounds $b_l$ and $b_h$ allows access to the memory region
% $[p+b_l,p+b_h)$.

\subsection{Checked Pointer Types}
%
\checkedc introduces three varieties of \emph{checked pointer}:
\begin{itemize}
\item \code{ptr<}$T$\code{>} types a pointer that is either null or
  points to a single object of type $T$.
\item \code{array_ptr<}$T$\code{>} types a pointer that is either null
  or points to an array of $T$ objects. The array width is defined
  by a \emph{bounds} expression, discussed below.
\item \code{nt_array_ptr<}$T$\code{>} is like
  \code{array_ptr<}$T$\code{>} except that the bounds expression
  defines the \emph{minimum} array width---additional objects may
  be available past the upper bound, up to a null terminator.
\end{itemize}
A bounds expression used with the latter two pointer types has three
forms:
\begin{itemize}
\item \code{count(}$e$\code{)} where $e$ defines the array's
  length. Thus, if pointer $p$ has bounds \code{count(n)} then the
  accessible memory is in the range $[p,p+$\code{n}$]$. Bounds
  expression $e$ must be side-effect free and may only refer to
  variables whose addresses are not taken, or adjacent \code{struct}
  fields.
\item \code{byte_count(}$e$\code{)} is like \code{count}, but
  expresses arithmetic using bytes, no objects; i.e.,
  \code{count(}$e$\code{)} used for \code{array_ptr<}$T$\code{>} is
  equivalent to \code{byte_count(}$e\times\texttt{sizeof}(T)$\code{)}
\item \code{bounds(}$e_l$,$e_h$\code{)} where $e_l$ and $e_h$ are
  pointers that bound the accessible region $[e_l,e_h)$ (the
  expressions are similarly restricted). Bounds
  \code{count(}$e$\code{)} is shorthand for
  \code{bounds(}$p, p + e$\code{)}. This most general form of bounds
  expression is useful for supporting pointer arithmetic.
\end{itemize}
  Dropping the bounds expression on an \code{nt_array_ptr} is equivalent
  to the bounds being \code{count(0)}.

The \checkedc compiler will instrument loads and stores of checked
pointers to confirm the pointer is non-null, and the access is within
the specified bounds. For pointers $p$ of type
\code{nt_array_ptr<}$T$\code{>}, such a check could spuriously fail if
the index is past $p$'s specified upper bound, but before the null
terminator. To address this problem, Checked C supports \emph{bounds
  widening}.
If $p$'s bounds expression is \code{bounds}$(e_l$,$e_h)$ a program may read from (but not
write to) $e_h$; when the compiler notices that a non-null character
is read at the upper bound, it will extend that bound to $e_h+1$.
% \liyi{solved: II.B.: LaTeX issue, your inline monospaced code snippets generate a space on
%   the following line (i.e. "a program may read from (but not write to)" is not
%   aligned with the left boundary of the paragraph).}

\subsection{Example}
\label{sec:example}
